---
title: "Climate Analysis"
author: "H. Wahid Khan, Vaibhav Kanodia"
date: "2025-03-09"
runtime: shiny
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Imports
We being by importing the necessary libraries.
```{r}
library(httr)
library(jsonlite)
library(ggplot2)
library(tidyr)
library(dplyr)
library(shiny)
```

## Data Acquisition
We make a helper function that:
1. Takes URL as input
2. Sends a GET request and parses the resulting JSON
3. Returns the parsed data
```{r}
get_data <- function(url) {
  encoded_url <- URLencode(url)
  response <- GET(encoded_url)

  if (http_status(response)$category == "Success") {
    json_data <- content(response, "text", encoding = "UTF-8")
    data <- fromJSON(json_data, flatten = TRUE)
    print(names(data))
  } else {
    print(paste("API request failed with status code:", http_status(response)$status_code))
    print(http_status(response)$reason) 
  }
  
  return(data)
}
```

We are using Open-Meteo's API for climate and air quality. These are test URLs that we gathered from their website frontend.
Later on, we will make these URLs and their parameters modifiable from the UI.
```{r}
climate_url <- "https://archive-api.open-meteo.com/v1/archive?latitude=23&longitude=25&start_date=1980-01-01&end_date=2025-01-01&hourly=apparent_temperature&daily=temperature_2m_max,temperature_2m_min,temperature_2m_mean,precipitation_sum"
airquality_url <- "https://air-quality-api.open-meteo.com/v1/air-quality?latitude=25&longitude=77&hourly=carbon_monoxide,carbon_dioxide,nitrogen_dioxide,sulphur_dioxide,methane&timezone=GMT&start_date=2013-01-01&end_date=2025-01-01&domains=cams_global"
```

Fetch the data from both the APIs.
```{r}
climate_data <- get_data(climate_url)
airquality_data <- get_data(airquality_url)
```

## Data Cleaning and Transformation
Here we take the previously acquired data and organize it into data frames. 
We have chosen to not handle NAs for air quality here, and instead handled it in the UI during rendering. By doing this we can make sure that all of the data is available to pick and choose from the UI.
```{r}
temp_max <- climate_data$daily$temperature_2m_max
temp_min <- climate_data$daily$temperature_2m_min
temp_mean <- climate_data$daily$temperature_2m_mean
date <- as.Date(climate_data$daily$time)

temp_df <- data.frame(
  date,
  temp_min,
  temp_mean,
  temp_max
)

co <- airquality_data$hourly$carbon_monoxide
ch4 <- airquality_data$hourly$methane
co2 <- airquality_data$hourly$carbon_dioxide
no2 <- airquality_data$hourly$nitrogen_dioxide
so2 <- airquality_data$hourly$sulphur_dioxide
date <- as.Date(airquality_data$hourly$time)

airquality_df <- data.frame(
  date,
  co,
  ch4,
  co2,
  no2,
  so2
)

airquality_df <- airquality_df[seq(1, nrow(airquality_df), 24), ]
```

Furthermore, the API only provides hourly data for air quality, so we are sampling every 24th data point to get daily data.

## Data Analysis
### Climate Analysis
For analyzing climate, we can take a look at the time series decomposition of the temperature, with a frequency of 365 (one year). This will help us visualize trends and seasonal patterns in the data better.

By using ggplot2, we are visualizing this decomposition.
```{r}
temp_ts <- ts(temp_df$temp_max, frequency = 365)
decomp_result <- stats::decompose(temp_ts)
trend_component <- decomp_result$trend
seasonal_component <- decomp_result$seasonal
remainder_component <- decomp_result$random

decomp_df <- data.frame(
  date = temp_df$date[1:length(temp_ts)],
  observed = as.numeric(temp_ts),
  trend = as.numeric(trend_component),
  seasonal = as.numeric(seasonal_component),
  remainder = as.numeric(remainder_component)
)

decomp_df <- decomp_df %>% 
  filter(!is.na(trend))

decomp_long <- decomp_df %>%
  pivot_longer(cols = c(observed, trend, seasonal, remainder),
               names_to = "component", 
               values_to = "value")

ggplot(decomp_long, aes(x = date, y = value)) +
  geom_line(color = "orange") +
  facet_wrap(~ component, scales = "free_y", ncol = 1) +
  labs(title = "Decomposition of Temperature Time Series",
       x = "Date",
       y = "Temperature (°C)") +
  theme_minimal() +
  theme(
    strip.background = element_rect(fill = "lightgray"),
    strip.text = element_text(face = "bold")
  )
```

We are using a bar chart to visualize the temperature as a time series dataset.
```{r}
temperature_data_long <- pivot_longer(
  temp_df,
  cols = c(temp_min, temp_max, temp_mean),
  names_to = "temperature_type",
  values_to = "temperature"
)
ggplot(temperature_data_long, aes(x = date, y = temperature, fill = temperature_type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.8) +
  labs(
    title = "Temperature Trends Over Time",
    x = "Date",
    y = "Temperature (°C)",
    fill = "Temperature Type"
  ) +
  scale_fill_manual(values = c("temp_min" = "blue", "temp_max" = "orange", "temp_mean" = "green")) +
  theme_minimal() +
  theme(
    panel.grid = element_line(color = "gray90"),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

### Air quality analysis
And for the air quality analysis, we are again using line plots to visualize the data as a time series dataset.
```{r}
airquality_long <- pivot_longer(
  airquality_df,
  cols = c(co, co2, ch4, no2, so2),
  names_to = "pollutant_type",
  values_to = "concentration"
)
ggplot(airquality_long, aes(x = date, y = concentration, color = pollutant_type, group = pollutant_type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Air Pollutant Concentrations Over Time",
    x = "Date",
    y = "Concentration",
    color = "Pollutant Type"
  ) +
  scale_color_manual(values = c("co" = "red", "co2" = "pink", ch4 = "purple", "no2" = "blue", "so2" = "green")) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Conclusion
