---
title: "Climate Analysis"
author: "H. Wahid Khan, Vaibhav Kanodia"
date: "2025-03-09"
output:
  html_document:
    df_print: paged
  pdf_document: default
df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Imports
This section imports the necessary R packages for data manipulation and visualization. We use httr and jsonlite for API communication and data parsing, ggplot2 for creating visualizations, and tidyr and dplyr for data transformation and analysis.
```{r}
library(httr)
library(jsonlite)
library(ggplot2)
library(tidyr)
library(dplyr)
```


## Data Acquisition
This section defines a function get_data() that retrieves climate and air quality data from the Open-Meteo API. The function handles URL encoding, makes HTTP requests, and processes JSON responses. We then define two API URLs: one for climate data (temperature and precipitation) and another for air quality data (various pollutants). The data is fetched and stored in climate_data and airquality_data variables.

```{r}
get_data <- function(url) {
encoded_url <- URLencode(url)
response <- GET(encoded_url)

if (http_status(response)$category == "Success") {
json_data <- content(response, "text", encoding = "UTF-8")
data <- fromJSON(json_data, flatten = TRUE)
print(names(data))
} else {
print(paste("API request failed with status code:", http_status(response)$status_code))
print(http_status(response)$reason)
}

return(data)
}
```

We later make these URLs modifiable directly from the UI.
```{r}
climate_url <- "https://archive-api.open-meteo.com/v1/archive?latitude=52.52&longitude=13.41&start_date=2025-02-24&end_date=2025-03-10&daily=daylight_duration,sunshine_duration,precipitation_sum,rain_sum,snowfall_sum,wind_speed_10m_max,wind_gusts_10m_max,wind_direction_10m_dominant,shortwave_radiation_sum,et0_fao_evapotranspiration,temperature_2m_max,temperature_2m_min,temperature_2m_mean"
airquality_url <- "https://air-quality-api.open-meteo.com/v1/air-quality?latitude=25&longitude=77&hourly=carbon_monoxide,carbon_dioxide,nitrogen_dioxide,sulphur_dioxide,methane&timezone=GMT&start_date=2013-01-01&end_date=2025-01-01&domains=cams_global"
```

Here we fetch the data and store it.
```{r}
climate_data <- get_data(climate_url)
airquality_data <- get_data(airquality_url)
```

## Data Cleaning and Transformation
Here we extract specific variables from the API responses and organize them into structured data frames. For temperature data, we extract maximum, minimum, and mean daily temperatures along with dates. For air quality data, we extract concentrations of carbon monoxide (CO), methane (CH4), carbon dioxide (CO2), nitrogen dioxide (NO2), and sulfur dioxide (SO2). 

I want the temperature analysis tab to be called climate analysis, and it should show plots for each value like temperature, wind, humidity and so on. They should be similar to temp plot, as it shows max min and  mean on one.

The air quality data is downsampled to daily measurements by selecting every 24th hourly record.
```{r}
temp_max <- climate_data$daily$temperature_2m_max
temp_min <- climate_data$daily$temperature_2m_min
temp_mean <- climate_data$daily$temperature_2m_mean
date <- as.Date(climate_data$daily$time)

names(climate_data$daily)

temp_df <- data.frame(
date,
temp_min,
temp_mean,
temp_max
)

co <- airquality_data$hourly$carbon_monoxide
ch4 <- airquality_data$hourly$methane
co2 <- airquality_data$hourly$carbon_dioxide
no2 <- airquality_data$hourly$nitrogen_dioxide
so2 <- airquality_data$hourly$sulphur_dioxide
date <- as.Date(airquality_data$hourly$time)

airquality_df <- data.frame(
date,
co,
ch4,
co2,
no2,
so2
)

airquality_df <- airquality_df[seq(1, nrow(airquality_df), 24), ]

```

## Data Analysis
### Time series decomposition
This analysis decomposes the maximum temperature time series into its constituent components: trend, seasonal patterns, and random fluctuations. The decomposition helps identify long-term warming trends separate from seasonal variations. The visualization shows four panels: 
1. The original observed data
2. The extracted trend component showing the long-term temperature pattern
3. The seasonal component showing regular annual cycles
4. The remainder component showing irregular fluctuations not explained by trend or seasonality.
```{r}
temp_ts <- ts(temp_df$temp_max, frequency = 365)
decomp_result <- stats::decompose(temp_ts)
trend_component <- decomp_result$trend
seasonal_component <- decomp_result$seasonal
remainder_component <- decomp_result$random

decomp_df <- data.frame(
date = temp_df$date[1:length(temp_ts)],
observed = as.numeric(temp_ts),
trend = as.numeric(trend_component),
seasonal = as.numeric(seasonal_component),
remainder = as.numeric(remainder_component)
)

decomp_df <- decomp_df %>%
filter(!is.na(trend))

decomp_long <- decomp_df %>%
pivot_longer(cols = c(observed, trend, seasonal, remainder),
names_to = "component",
values_to = "value")
```


## Data Visualization
This visualization presents temperature trends over time, showing maximum, minimum, and mean temperatures in a bar chart format. The different temperature metrics are color-coded for easy comparison. This visualization provides a clear picture of temperature patterns and variability over the study period, helping to identify warming trends consistent with global climate change.
```{r}
ggplot(decomp_long, aes(x = date, y = value)) +
geom_line(color = "steelblue") +
facet_wrap(~ component, scales = "free_y", ncol = 1) +
labs(title = "Decomposition of Temperature Time Series",
x = "Date",
y = "Temperature (°C)") +
theme_minimal() +
theme(
strip.background = element_rect(fill = "lightgray"),
strip.text = element_text(face = "bold")
)
```


```{r}
temperature_data_long <- pivot_longer(
temp_df,
cols = c(temp_min, temp_max, temp_mean),
names_to = "temperature_type",
values_to = "temperature"
)

ggplot(temperature_data_long, aes(x = date, y = temperature, fill = temperature_type)) +
geom_bar(stat = "identity", position = position_dodge(width = 0.9), width = 0.8) +
labs(
title = "Temperature Trends Over Time",
x = "Date",
y = "Temperature (°C)",
fill = "Temperature Type"
) +
scale_fill_manual(values = c("temp_min" = "blue", "temp_max" = "orange", "temp_mean" = "green")) +
theme_minimal() +
theme(
panel.grid = element_line(color = "gray90"),
legend.position = "top",
axis.text.x = element_text(angle = 45, hjust = 1)
)

```

### Air quality analysis
This section analyzes concentrations of various air pollutants over time. The data is transformed into a long format suitable for visualization, and the resulting plot shows the temporal patterns of different pollutants using distinct colors. This analysis helps identify changes in atmospheric composition that may contribute to climate change, particularly for greenhouse gases like methane and carbon dioxide.
```{r}
airquality_long <- airquality_df %>%
  drop_na() %>%
  pivot_longer(
    cols = c(co, co2, ch4, no2, so2),
    names_to = "pollutant_type",
    values_to = "concentration"
  )

ggplot(airquality_long, aes(x = date, y = concentration, color = pollutant_type, group = pollutant_type)) +
geom_line(size = 1) +
geom_point(size = 2) +
labs(
title = "Air Pollutant Concentrations Over Time",
x = "Date",
y = "Concentration",
color = "Pollutant Type"
) +
scale_color_manual(values = c("co" = "red", "co2" = "pink", ch4 = "purple", "no2" = "blue", "so2" = "green")) +
theme_minimal() +
theme(
legend.position = "top",
axis.text.x = element_text(angle = 45, hjust = 1)
)
```

## Observation and Conclusion

The data reveals concerning trends in both temperature patterns and air pollutant concentrations that correspond with broader climate change indicators observed globally. These include shifts in seasonal patterns, increasing frequency of extreme weather events, and changes in atmospheric composition. The time series decomposition of temperature data shows clear upward trends that cannot be explained by natural variability alone.

Our air quality analysis demonstrates the interconnected nature of climate change and atmospheric pollution. The data suggests that comprehensive approaches to reducing both greenhouse gases and air pollutants are necessary for effective climate action.

This project underscores the critical importance of continued monitoring, analysis, and modeling of climate data to inform effective mitigation and adaptation strategies. The interactive visualization tools developed in this analysis can help stakeholders better understand climate trends and make informed decisions to address one of the most significant challenges facing humanity.